---
title: "Introductory chapter questions"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Topic 1

### Exercise 

Distance sampling is an extension of plot sampling, so there are some similarities between the two.  These questions ask you to compare between the two methods.

```{r plotsamp}
quiz(
  question("What would be the appropriate key function to model data collected using plot sampling?",
    answer("uniform", correct = TRUE),
    answer("half normal"),
    answer("hazard rate"),
    answer("negative exponential")
  ),
  question("Why should there be uncertainty in density estimates generated by plot sampling when there is no detection function to fit?",
    answer("Because only a portion of the study area is sampled; inference must be drawn to portions of the study area not visited.", correct = TRUE),
    answer(""),
    answer(""),
    answer("")
  )
)
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

